---
title: "C-02 notebook — R"
subtitle: "A regimen approved without a trial"
execute:
  freeze: auto
---

The same argument in R. Base R only — the two-compartment steady state has a
closed form, so no solver and no packages are needed.

::: {.callout-warning appearance="simple"}
## About these parameters
Not Merck's estimates. Values chosen from inside the published range for
anti-PD-1 antibodies — clearance 0.179–0.290 L/day, central volume 2.98–4.46 L
(Shang et al., [DOI](https://doi.org/10.3389/fimmu.2022.871372)). The structure
of the argument is what transfers, not the numbers.
:::

```{r}
CL <- 0.22; V1 <- 3.5; Q <- 0.40; V2 <- 3.0     # L/day, L, L/day, L
k10 <- CL / V1; k12 <- Q / V1; k21 <- Q / V2
WT  <- 75

s     <- k10 + k12 + k21
alpha <- 0.5 * (s + sqrt(s^2 - 4 * k10 * k21))
beta  <- 0.5 * (s - sqrt(s^2 - 4 * k10 * k21))

# Exact steady-state concentration for repeated IV bolus, two compartments.
conc_ss <- function(t, dose, tau) {
  A <- (dose / V1) * (alpha - k21) / (alpha - beta)
  B <- (dose / V1) * (k21 - beta)  / (alpha - beta)
  A * exp(-alpha * t) / (1 - exp(-alpha * tau)) +
  B * exp(-beta  * t) / (1 - exp(-beta  * tau))
}

metrics_ss <- function(dose, tau, n = 4001) {
  t <- seq(0, tau, length.out = n)
  c <- conc_ss(t, dose, tau)
  list(Cavg = dose / (CL * tau),   # exact: AUC over an interval = Dose / CL
       Cmin = c[n], Cmax = c[1], per_day = dose / tau, t = t, c = c)
}

cat(sprintf("alpha = %.4f /day  -> distribution half-life %.2f days\n",
            alpha, log(2) / alpha))
cat(sprintf("beta  = %.4f /day  -> terminal half-life     %.1f days\n",
            beta, log(2) / beta))
```

## The three regimens

```{r}
regimens <- list("200 mg Q3W"   = c(200, 21),
                 "400 mg Q6W"   = c(400, 42),
                 "10 mg/kg Q2W" = c(10 * WT, 14))

m <- lapply(regimens, function(r) metrics_ss(r[1], r[2]))

cat(sprintf("%14s %8s %8s %8s %8s\n",
            "regimen", "mg/day", "Cavg", "Cmin", "Cmax"))
cat(strrep("-", 50), "\n")
for (nm in names(m))
  cat(sprintf("%14s %8.1f %8.2f %8.2f %8.2f\n",
              nm, m[[nm]]$per_day, m[[nm]]$Cavg, m[[nm]]$Cmin, m[[nm]]$Cmax))
cat("\nconcentrations in mg/L, at steady state\n")
```

## Leg 1 — Cavg matches by algebra

```{r}
q3w <- m[["200 mg Q3W"]]; q6w <- m[["400 mg Q6W"]]
ratio <- q6w$Cavg / q3w$Cavg
cat(sprintf("Cavg ratio Q6W / Q3W = %.6f\n", ratio))

# Cavg = Dose / (CL * tau). 400/42 == 200/21, so this is exactly 1 and the
# assertion cannot fail -- which is precisely the point being made.
stopifnot(abs(ratio - 1) < 1e-12)
cat("\nExactly 1, not approximately. This leg is arithmetic and tells you\n",
    "nothing about the drug.\n")
```

## Legs 2 and 3 — the trough and the peak

```{r}
ceiling_cmax <- m[["10 mg/kg Q2W"]]$Cmax

cat(sprintf("Cmin  Q3W %.2f   Q6W %.2f   (%.1f%% lower)\n",
            q3w$Cmin, q6w$Cmin, 100 * (1 - q6w$Cmin / q3w$Cmin)))
cat(sprintf("Cmax  Q6W %.2f   ceiling %.2f   (%.0f%% below)\n",
            q6w$Cmax, ceiling_cmax, 100 * (1 - q6w$Cmax / ceiling_cmax)))

stopifnot(q6w$Cmin < q3w$Cmin,          # trough must fall
          q6w$Cmax > q3w$Cmax,          # peak must rise
          q6w$Cmax < ceiling_cmax)      # and stay under what was tolerated
cat("\nThe trough falls and needs a target-saturation argument.\n",
    "The peak rises but stays below one patients have already received.\n")
```

## The picture

```{r}
#| fig-width: 9
#| fig-height: 6.2
tile <- function(nm, horizon = 84) {
  d <- regimens[[nm]]; tau <- d[2]; x <- m[[nm]]
  reps <- ceiling(horizon / tau)
  tt <- unlist(lapply(0:(reps - 1), function(i) x$t + i * tau))
  cc <- rep(x$c, reps)
  k <- tt <= horizon
  list(t = tt[k], c = cc[k])
}
COL <- c("200 mg Q3W" = "#2A5DB0", "400 mg Q6W" = "#A32C42",
         "10 mg/kg Q2W" = "#98A4AC")

op <- par(mfrow = c(2, 1), mar = c(2.2, 4, 2.4, 1), cex = 0.85)

# Leg 2: the comparison, at a scale where the trough difference is visible
plot(NA, xlim = c(0, 84), ylim = c(0, 145), xlab = "", ylab = "conc. (mg/L)",
     main = "", bty = "n", las = 1)
title("Leg 2 — the trough falls, and must be defended", adj = 0, cex.main = 1)
for (nm in c("200 mg Q3W", "400 mg Q6W")) {
  z <- tile(nm); lines(z$t, z$c, col = COL[nm], lwd = 2)
}
abline(h = q3w$Cmin, col = "#66737C", lty = 3)
legend("topright", names(COL)[1:2], col = COL[1:2], lwd = 2, bty = "n")

# Leg 3: the same curves against the ceiling
par(mar = c(4, 4, 2.4, 1))
plot(NA, xlim = c(0, 84), ylim = c(0, 410), xlab = "days at steady state",
     ylab = "conc. (mg/L)", main = "", bty = "n", las = 1)
title("Leg 3 — the new peak sits below one already tolerated",
      adj = 0, cex.main = 1)
for (nm in c("10 mg/kg Q2W", "200 mg Q3W", "400 mg Q6W")) {
  z <- tile(nm)
  lines(z$t, z$c, col = COL[nm], lwd = if (nm == "10 mg/kg Q2W") 1.3 else 2,
        lty = if (nm == "10 mg/kg Q2W") 2 else 1)
}
abline(h = ceiling_cmax, col = "#66737C", lty = 3)
legend("topright", names(COL), col = COL, lwd = 2, lty = c(2, 1, 1), bty = "n")
par(op)
```

## The failure mode

```{r}
q12 <- metrics_ss(800, 84)
cat(sprintf("800 mg Q12W:  Cavg %.2f  (ratio %.6f)\n",
            q12$Cavg, q12$Cavg / q3w$Cavg))
cat(sprintf("              Cmin %.2f  vs Q3W %.2f  (%.0f%% lower)\n",
            q12$Cmin, q3w$Cmin, 100 * (1 - q12$Cmin / q3w$Cmin)))
cat(sprintf("              Cmax %.2f  vs ceiling %.2f\n",
            q12$Cmax, ceiling_cmax))
cat("\nCavg matches exactly, and everything that matters has left the range\n",
    "the exposure-response data covers. A table reporting only Cavg would\n",
    "call this a success.\n")
```

## What you should take away

Cavg matching is guaranteed by arithmetic. The argument lives in the trough and
the peak, and only inside the exposure range the programme actually studied.
