---
title: "C-03 notebook — R"
subtitle: "A flat exposure–response"
execute:
  freeze: auto
---

The same experiment as the Python notebook: a trial in which exposure has
exactly no effect on outcome, an exposure–response analysis that reports a large
one anyway, and the diagnostic that catches it.

Base R only — no packages. The random number generator differs from NumPy's, so
the numbers here will not match the Python notebook digit for digit. Every
conclusion does, and that is the part worth checking.

```{r}
INK <- "#3A464E"; MUTED <- "#66737C"; GRID <- "#DFE5E8"
DRUG <- "#2A5DB0"; PROGNOSIS <- "#A32C42"
```

## The generative model

Antibody clearance rises with catabolic burden — the sicker the patient, the
faster the drug goes away. Sicker patients also die sooner, for reasons that
have nothing to do with the drug. Those two facts alone are enough.

```{r}
simulate <- function(seed = 20260910, n = 1200, theta_cl = 0.40, beta_s = 0.55,
                     true_te = 0.70, beta_exp = 0.00, proxy_sd = 0.50) {
  ## theta_cl  severity -> clearance   (catabolism)
  ## beta_s    severity -> hazard      (prognosis)
  ## true_te   the drug's real effect  (hazard ratio vs no drug)
  ## beta_exp  exposure -> hazard      (THE ANSWER; 0.00 means none)
  ## proxy_sd  how badly the recorded covariate measures true severity
  set.seed(seed)
  CL0 <- 0.25; TAU <- 14; DOSE <- 240; H0 <- 0.020; TMAX <- 36

  arm <- function(treated) {
    S    <- rnorm(n)                                  # true severity, never observed
    X    <- S + rnorm(n, 0, proxy_sd)                 # what the case report form records
    CL   <- CL0 * exp(theta_cl * S + rnorm(n, 0, 0.25))
    cavg <- DOSE / (CL * TAU)                         # exposure falls as clearance rises
    h    <- H0 *
            exp(beta_s * S) *                         # prognosis drives the hazard
            (if (treated) true_te else 1) *           # the drug works, by a fixed amount
            exp(beta_exp * log(cavg / median(cavg)))
    t <- rexp(n, rate = h)
    list(S = S, X = X, CL = CL, cavg = cavg,
         time = pmin(t, TMAX), event = t <= TMAX)
  }
  list(treated = arm(TRUE), control = arm(FALSE))
}

sim <- simulate()
treated <- sim$treated; control <- sim$control
cat(sprintf("treated %d, control %d\n", length(treated$time), length(control$time)))
cat(sprintf("events:  %d vs %d\n", sum(treated$event), sum(control$event)))
```

Exposure enters the hazard through `beta_exp` and nothing else. At `0.00` the
term is `exp(0 * log(...)) = 1` for every patient, so exposure is arithmetically
absent from every survival time in the dataset.

```{r}
## The parameter has to be live, or "we set it to zero" proves nothing.
same    <- simulate(beta_exp =  0.00)$treated
with_er <- simulate(beta_exp = -0.50)$treated
stopifnot(identical(same$cavg, with_er$cavg))            # same patients, same exposures
stopifnot(!identical(same$time, with_er$time))           # but different outcomes
cat("beta_exp = 0.00  ->  exposure multiplier is exactly 1 for all",
    length(same$time), "patients\n")
cat("true exposure effect per e-fold of Cavg: HR = 1.000000 (by construction)\n")
```

## The hazard ratio, without an optimiser

For exponential survival the hazard rate is events divided by person-time, and a
hazard ratio is one such rate divided by another. No `coxph`, no solver — and
the arithmetic *is* the definition.

```{r}
hazard_ratio <- function(dA, mA, dB, mB) {
  eA <- sum(dA$event[mA]); tA <- sum(dA$time[mA])
  eB <- sum(dB$event[mB]); tB <- sum(dB$time[mB])
  hr <- (eA / tA) / (eB / tB)
  se <- sqrt(1 / eA + 1 / eB)                            # standard error of log HR
  c(hr = hr, lo = hr * exp(-1.96 * se), hi = hr * exp(1.96 * se))
}

quartiles <- function(d, key = "cavg") {
  v <- d[[key]]; q <- unname(quantile(v, c(.25, .50, .75)))
  list(v <= q[1], v > q[1] & v <= q[2], v > q[2] & v <= q[3], v > q[3])
}

Q <- quartiles(treated)
cat(sprintf("%4s%11s%15s%8s%9s\n", "", "mean Cavg", "mean severity",
            "events", "hazard"))
for (i in 1:4) {
  m <- Q[[i]]; e <- sum(treated$event[m]); pt <- sum(treated$time[m])
  cat(sprintf("  Q%d%11.1f%+15.2f%8d%9.4f\n", i, mean(treated$cavg[m]),
              mean(treated$S[m]), e, e / pt))
}
```

The exposure quartiles differ in exposure. They also differ, by more than a
standard deviation end to end, in how sick the patients were before anyone gave
them anything.

## The same picture, twice

```{r}
#| fig-width: 9
#| fig-height: 3.9
haz <- sapply(Q, function(m) sum(treated$event[m]) / sum(treated$time[m]))
sev <- sapply(Q, function(m) mean(treated$S[m]))
lab <- c("Q1", "Q2", "Q3", "Q4")

op <- par(mfrow = c(1, 2), mar = c(4.2, 4.4, 3.4, 1), col.axis = INK,
          col.lab = INK, fg = MUTED, las = 1, cex.axis = 0.85)

## No grid on these two panels. Drawing one behind base-R bars means calling
## barplot twice with matching `space`, and a silent mismatch there shifts the
## bars off their own axis. The value labels carry the numbers instead.
b <- barplot(haz, names.arg = lab, col = DRUG, border = NA, space = 0.6,
             ylim = c(0, max(haz) * 1.22), ylab = "events per patient-month")
text(b, haz, sprintf("%.4f", haz), pos = 3, cex = 0.8, col = INK)
title("What the analysis reports\nhazard rate by exposure quartile",
      adj = 0, cex.main = 0.95, font.main = 1, col.main = INK)

b2 <- barplot(sev, names.arg = lab, col = PROGNOSIS, border = NA, space = 0.6,
              ylim = c(min(sev) * 1.35, max(sev) * 1.35),
              ylab = "severity score (0 = average)")
abline(h = 0, col = MUTED)
text(b2, sev, sprintf("%+.2f", sev), pos = ifelse(sev > 0, 3, 1),
     cex = 0.8, col = INK)
title("What the quartiles actually differ in\nmean baseline severity",
      adj = 0, cex.main = 0.95, font.main = 1, col.main = INK)
par(op)
```

Two gradients, one cause. Sorting patients by exposure sorted them by prognosis,
because clearance is a symptom of the disease.

## 1. The analysis anyone would run

```{r}
naive <- hazard_ratio(treated, Q[[4]], treated, Q[[1]])
cat("highest vs lowest exposure quartile\n")
cat(sprintf("  HR = %.2f   95%% CI %.2f to %.2f\n", naive[1], naive[2], naive[3]))
cat("  truth       1.00\n")
```

A large reduction in hazard, with a confidence interval nowhere near 1, from a
drug whose exposure does nothing. This is not noise and it will not go away with
a larger trial — more patients would tighten that interval around the wrong
number.

## 2. Adjusting for prognosis does not fix it

```{r}
xq <- unname(quantile(treated$X, c(.2, .4, .6, .8)))
strata <- list(treated$X <= xq[1],
               treated$X > xq[1] & treated$X <= xq[2],
               treated$X > xq[2] & treated$X <= xq[3],
               treated$X > xq[3] & treated$X <= xq[4],
               treated$X > xq[4])
num <- 0; den <- 0
for (s in strata) {
  eA <- sum(treated$event[s & Q[[4]]]); tA <- sum(treated$time[s & Q[[4]]])
  eB <- sum(treated$event[s & Q[[1]]]); tB <- sum(treated$time[s & Q[[1]]])
  if (eA > 0 && eB > 0) { num <- num + eA / tA; den <- den + eB / tB }
}
cat(sprintf("  HR stratified on the recorded covariate = %.2f\n", num / den))
cat(sprintf("  unadjusted                              = %.2f\n", naive[1]))
cat        ("  truth                                   = 1.00\n")
```

Part of the bias, and part is not enough. The covariate you recorded is a noisy
reading of the thing that actually drives both clearance and death, and
adjusting for a noisy proxy removes only the part of the confounding the proxy
managed to capture.

## 3. The diagnostic that works

Stop asking whether high exposure beats low exposure inside the treated arm. Ask
instead whether the **lowest**-exposure patients still beat comparable patients
who never got the drug.

```{r}
match_controls <- function(treated, control, mask, on = "X") {
  pool <- seq_along(control[[on]]); chosen <- integer(0)
  for (i in which(mask)) {
    k <- which.min(abs(control[[on]][pool] - treated[[on]][i]))
    chosen <- c(chosen, pool[k]); pool <- pool[-k]
  }
  m <- rep(FALSE, length(control[[on]])); m[chosen] <- TRUE; m
}

mX <- match_controls(treated, control, Q[[1]], "X")
matched <- hazard_ratio(treated, Q[[1]], control, mX)
cat(sprintf("matched on the recorded covariate:  treated Q1 mean %+.2f vs controls %+.2f\n",
            mean(treated$X[Q[[1]]]), mean(control$X[mX])))
cat(sprintf("  HR (lowest-exposure treated vs matched controls) = %.2f   95%% CI %.2f to %.2f\n",
            matched[1], matched[2], matched[3]))
cat        ("  the drug's true effect                           = 0.70\n")

mS <- match_controls(treated, control, Q[[1]], "S")
oracle <- hazard_ratio(treated, Q[[1]], control, mS)
cat(sprintf("  matched on TRUE severity (not available in life) = %.2f   95%% CI %.2f to %.2f\n",
            oracle[1], oracle[2], oracle[3]))
```

The patients with the least drug exposure still got most of the drug's benefit.
That is incompatible with a real exposure–response and compatible with the
truth: the drug works, and the dose is on a plateau.

Matching on the truth recovers something close to 0.70. Matching on what was
written down lands short of it. The gap is the price of measuring prognosis
imperfectly, and in a real study nobody can show you that gap.

```{r}
#| fig-width: 9
#| fig-height: 2.9
rows    <- list(naive, matched, oracle)
names_  <- c("naive: Q4 vs Q1 exposure", "matched on recorded covariate",
             "matched on true severity")
targets <- c(1.00, 0.70, 0.70)
tnames  <- c("no exposure effect", "the drug's effect", "the drug's effect")

op <- par(mar = c(4.2, 12, 1, 12), col.axis = INK, col.lab = INK, fg = MUTED,
          las = 1, cex.axis = 0.85)
plot(NA, xlim = c(0.2, 2), ylim = c(0.4, 3.6), log = "x", xaxt = "n",
     yaxt = "n", xlab = "", ylab = "", bty = "n")
abline(v = c(0.25, 0.5, 1, 2), col = GRID)
abline(v = 1, col = MUTED, lty = 3)
axis(1, at = c(0.25, 0.5, 1, 2), labels = c("0.25", "0.5", "1.0", "2.0"))
mtext("hazard ratio (log scale) — filled circle: estimate, open diamond: truth",
      side = 1, line = 2.6, cex = 0.85, col = INK)
for (i in seq_along(rows)) {
  y <- length(rows) + 1 - i
  segments(rows[[i]][2], y, rows[[i]][3], y, col = DRUG, lwd = 3)
  points(rows[[i]][1], y, pch = 21, bg = DRUG, col = "white", cex = 1.5, lwd = 1.5)
  points(targets[i], y, pch = 5, col = PROGNOSIS, cex = 1.5, lwd = 2)
  text(rows[[i]][1], y + 0.26, sprintf("%.2f", rows[[i]][1]), cex = 0.8, col = INK)
  mtext(names_[i], side = 2, at = y, line = 0.5, cex = 0.85, col = INK)
  mtext(sprintf("target %.2f — %s", targets[i], tnames[i]), side = 4, at = y,
        line = 0.5, cex = 0.85, col = MUTED, adj = 0)
}
par(op)
```

## The tests, and the sabotage that makes them tests

An assertion that passes whether or not the thing it guards is broken is
decoration. Each check below is paired with a deliberately broken run that must
make it fail.

```{r}
flat      <- simulate(theta_cl = 0.0)
flatQ     <- quartiles(flat$treated)
flat_naive <- hazard_ratio(flat$treated, flatQ[[4]], flat$treated, flatQ[[1]])

perf   <- simulate(proxy_sd = 0.01)
perfQ  <- quartiles(perf$treated)
perf_matched <- hazard_ratio(perf$treated, perfQ[[1]], perf$control,
                             match_controls(perf$treated, perf$control,
                                            perfQ[[1]], "X"))

checks <- list(
  c("a zero exposure effect is reported as a large one",
    naive[3] < 0.75),
  c("and it is not a confidence problem — the interval excludes 1",
    naive[3] < 1.0),
  c("removing the clearance-prognosis link removes the whole signal",
    flat_naive[2] < 1.0 && flat_naive[3] > 1.0),
  c("case-matching moves the estimate most of the way back",
    matched[1] > 2 * naive[1]),
  c("recording prognosis better closes the residual gap",
    abs(perf_matched[1] - oracle[1]) < abs(matched[1] - oracle[1]))
)
for (ch in checks) cat(sprintf("  %s  %s\n",
                               if (as.logical(ch[2])) "PASS" else "FAIL", ch[1]))
stopifnot(all(sapply(checks, function(ch) as.logical(ch[2]))))

cat(sprintf("\n  sabotage 1 (theta_cl = 0):  HR %.2f (%.2f-%.2f)  <- the bias is gone\n",
            flat_naive[1], flat_naive[2], flat_naive[3]))
cat(sprintf("  sabotage 2 (perfect covariate): matched HR %.2f vs %.2f with a noisy one\n",
            perf_matched[1], matched[1]))
```

Sabotage 1 is the important one. Set the clearance–prognosis link to zero and the
spurious effect vanishes — which is the proof that the link, and nothing else in
this simulation, produced it.

## What you should take away

An exposure–response analysis inside a single treatment arm is an observational
study wearing a pharmacokinetic costume. Randomisation assigned the dose; it did
not assign the exposure, and whatever determines the exposure is free to
determine the outcome too.
