---
title: "P-04 notebook — R"
subtitle: "Binding, quantified"
execute:
  freeze: auto
---

```{r setup}
#| message: false
library(deSolve)
library(ggplot2)
theme_set(theme_minimal(base_size = 11))

nM   <- 1e-9
Kd   <- 120 * nM        # dissociation constant, M
kon  <- 1e5             # association rate, 1/(M*s)
koff <- kon * Kd        # dissociation rate, 1/s
```

## 1. The occupancy curve

```{r}
occupancy <- function(conc, Kd) conc / (conc + Kd)

for (m in c(0.1, 1, 9, 99)) {
  cat(sprintf("  free drug = %5.1f x Kd   ->  occupancy %5.1f%%\n",
              m, 100 * occupancy(m * Kd, Kd)))
}
```

Half occupancy costs one K~d~. Ninety percent costs nine. Ninety-nine percent
costs ninety-nine.

```{r}
#| fig-height: 3.6
conc <- 10^seq(-10, -5, length.out = 400)
marks <- data.frame(f = c(.1, .5, .9))
marks$conc <- Kd * marks$f / (1 - marks$f)

ggplot(data.frame(conc = conc, occ = 100 * occupancy(conc, Kd)),
       aes(conc / nM, occ)) +
  geom_line(linewidth = 0.9, colour = "#0F6E63") +
  geom_point(data = marks, aes(conc / nM, 100 * f), colour = "#8E6410", size = 2) +
  scale_x_log10() +
  labs(x = "Free drug concentration (nM, log scale)",
       y = "Target occupancy (%)", title = "Langmuir occupancy, Kd = 120 nM")
```

```{r}
c10 <- Kd * .1 / .9; c90 <- Kd * .9 / .1
cat(sprintf("10%% occupancy at %6.1f nM\n", c10 / nM))
cat(sprintf("90%% occupancy at %6.1f nM\n", c90 / nM))
cat(sprintf("ratio          : %.0f-fold   <- always 81 for simple binding\n", c90 / c10))
```

**Memorise the 81.** Simple 1:1 binding always spans 10%–90% over an 81-fold
concentration range. Anything steeper means cooperativity, amplification or a
threshold — see `P-14`.

## 2. Rates, not just the ratio

```{r}
cat(sprintf("Kd                     : %.0f nM\n", Kd / nM))
cat(sprintf("koff                   : %.2e 1/s\n", koff))
cat(sprintf("residence time         : %.1f s  (%.1f min)\n", 1 / koff, 1 / koff / 60))
cat(sprintf("dissociation half-life : %.1f s\n", log(2) / koff))
```

### How fast does binding reach equilibrium?

```{r}
approach <- function(C, times) {
  rhs <- function(t, y, p) list(kon * C * (1 - y[1]) - koff * y[1])
  as.data.frame(ode(c(DR = 0), times, rhs, parms = NULL,
                    method = "lsoda", rtol = 1e-10, atol = 1e-14))
}

times <- seq(0, 300, by = 0.5)
res <- do.call(rbind, lapply(c(1, 10), function(m) {
  C <- m * Kd
  d <- approach(C, times); d$mult <- factor(paste(m, "x Kd"))
  tau <- 1 / (kon * C + koff)
  cat(sprintf("  C = %2d x Kd  equilibrium %5.1f%%   tau = %5.1f s   90%% of equilibrium by %5.0f s\n",
              m, 100 * occupancy(C, Kd), tau, 2.303 * tau))
  d
}))
```

```{r}
#| fig-height: 3.6
ggplot(res, aes(time, 100 * DR, colour = mult)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c("1 x Kd" = "#5FBCAC", "10 x Kd" = "#0B4F47")) +
  labs(x = "Time (s)", y = "Occupancy (%)", colour = "free drug",
       title = "Approach to binding equilibrium")
```

Higher concentration equilibrates **faster** — the on-rate term grows with
concentration, the off-rate does not. Both are seconds to minutes, against
plasma concentrations moving over hours.

## 3. What an IC₅₀ is measuring

```{r}
cheng_prusoff <- function(IC50, L, Kd_L) IC50 / (1 + L / Kd_L)

IC50 <- 250 * nM
cat("same drug, same target, three assay setups:\n\n")
for (r in c(0.1, 1, 10)) {
  cat(sprintf("  competing ligand at %4g x its Kd  ->  Ki = %5.0f nM\n",
              r, cheng_prusoff(IC50, r * Kd, Kd) / nM))
}
```

A tenfold spread in apparent potency, decided by what was in the well.

## 4. Ozanib against a 120 nM target

```{r}
MW <- 465; FU <- 0.08
ng_ml_to_M <- function(x) x / MW * 1e-6

for (nm in c("steady-state trough", "steady-state peak")) {
  ng <- if (nm == "steady-state trough") 51.9 else 560.7
  total <- ng_ml_to_M(ng); free <- total * FU
  cat(sprintf("  %-20s total %7.1f nM   free %6.1f nM   occupancy %5.1f%%\n",
              nm, total / nM, free / nM, 100 * occupancy(free, Kd)))
}
```

Occupancy swings from **7% to 45%** across a dosing interval and never
approaches saturation.

## 5. The assumption that quietly fails

```{r}
occupancy_exact <- function(D_tot, R_tot, Kd) {
  b <- D_tot + R_tot + Kd
  DR <- 0.5 * (b - sqrt(b^2 - 4 * D_tot * R_tot))
  DR / R_tot
}

cat(sprintf("%10s  %8s  %8s   target = 100 nM\n", "drug (nM)", "simple", "exact"))
for (D in c(10, 30, 100, 300, 1000, 3000)) {
  s <- 100 * occupancy(D * nM, Kd)
  e <- 100 * occupancy_exact(D * nM, 100 * nM, Kd)
  cat(sprintf("%10d  %7.1f%%  %7.1f%%%s\n", D, s, e,
              if (s - e > 5) sprintf("   <- overstated by %.0f points", s - e) else ""))
}
```

At low concentrations the simple formula is badly wrong: much of the drug is on
the target rather than free. Every therapeutic antibody against a circulating
target lives here — and the bound complex is cleared, so the target eliminates
the drug. That is target-mediated disposition, built in `S1`.

```{r}
#| fig-height: 3.6
D <- 10^seq(0, 4, length.out = 200) * nM
data.frame(D = rep(D, 2),
           occ = c(100 * occupancy(D, Kd), 100 * occupancy_exact(D, 100 * nM, Kd)),
           model = rep(c("simple (drug in excess)", "exact (target = 100 nM)"), each = length(D))) |>
  ggplot(aes(D / nM, occ, colour = model, linetype = model)) +
  geom_line(linewidth = 0.9) + scale_x_log10() +
  scale_colour_manual(values = c("simple (drug in excess)" = "#0F6E63",
                                 "exact (target = 100 nM)" = "#8E6410")) +
  labs(x = "Total drug concentration (nM, log scale)", y = "Target occupancy (%)",
       colour = NULL, linetype = NULL,
       title = "When target concentration is not negligible") +
  theme(legend.position = "top")
```

## What to try next

1. Set `kon <- 1e3` keeping `Kd` fixed. How long is residence time now, and how
   does equilibration compare with a dosing interval?
2. At what target concentration does the simple formula stay within 5% of the
   exact one everywhere? That is your rule for when the shortcut is safe.
3. Take the day-14 Ozanib profile from `S0-01` and plot occupancy across the
   dosing interval. What fraction of 24 h is occupancy above 25%?
