---
title: "P-05 notebook — R"
subtitle: "Agonists, antagonists and everything between"
execute:
  freeze: auto
---

```{r setup}
#| message: false
library(ggplot2)
theme_set(theme_minimal(base_size = 11))

nM <- 1e-9
KA <- 120 * nM       # agonist affinity — the same target as P-04
Em <- 100            # maximum the system can produce, arbitrary units

TEAL <- "#0F6E63"; RUST <- "#8E6410"; SLATE <- "#4C5B54"
```

## 1. The operational model

Occupancy drives a saturable transduction step. One parameter, `tau`, carries
both the efficacy of the drug and the responsiveness of the tissue.

```{r}
# Black & Leff operational model. KA and Em are taken from the setup chunk.
response  <- function(A, tau = 1, KA. = KA, Em. = Em) Em. * tau * A / (A * (tau + 1) + KA.)
occupancy <- function(A, KA. = KA)                    A / (A + KA.)
emax_obs  <- function(tau, Em. = Em)                  Em. * tau / (tau + 1)
ec50_obs  <- function(tau, KA. = KA)                  KA. / (tau + 1)
```

The two closed-form results the lesson claims. Confirm them numerically rather
than taking them on trust — solve for the half-maximal point on a fine grid and
compare.

```{r}
A <- 10^seq(-11, -4, length.out = 20001)

cat(sprintf("%6s %9s %11s %12s %9s %10s\n",
            "tau", "Emax obs", "EC50 (nM)", "KA/(1+tau)", "occ@EC50", "1/(tau+2)"))
for (tau in c(0.5, 1, 3, 9, 99)) {
  E        <- response(A, tau = tau)
  half     <- emax_obs(tau) / 2
  ec50_num <- A[which.min(abs(E - half))]
  ec50_ana <- ec50_obs(tau)
  stopifnot(isTRUE(all.equal(ec50_num, ec50_ana, tolerance = 2e-3)))
  occ <- occupancy(ec50_ana)
  stopifnot(isTRUE(all.equal(occ, 1 / (tau + 2))))
  cat(sprintf("%6g %9.1f %11.2f %12.2f %8.1f%% %9.1f%%\n",
              tau, emax_obs(tau), ec50_num / nM, ec50_ana / nM,
              100 * occ, 100 / (tau + 2)))
}
```

Both identities hold exactly. The third column is what an experiment reports;
the affinity is fixed at 120 nM throughout and never appears in it.

## 2. Receptor reserve

The binding curve does not move. The response curve does.

```{r}
#| fig-height: 4.2
taus <- c(0.5, 3, 9, 99)
cols <- c("#C9A227", RUST, "#2E8B7A", TEAL)

curves <- do.call(rbind, lapply(seq_along(taus), function(i) {
  data.frame(A = A, y = response(A, tau = taus[i]),
             lab = paste0("response, tau = ", taus[i]))
}))
occ_df <- data.frame(A = A, y = 100 * occupancy(A), lab = "occupancy (fixed)")
pts <- data.frame(A = ec50_obs(taus), y = emax_obs(taus) / 2,
                  lab = paste0("response, tau = ", taus))

ggplot(curves, aes(A / nM, y, colour = lab)) +
  geom_line(data = occ_df, aes(A / nM, y), colour = SLATE,
            linetype = "dashed", linewidth = 0.9, inherit.aes = FALSE) +
  geom_line(linewidth = 0.85) +
  geom_point(data = pts, size = 1.9) +
  geom_vline(xintercept = KA / nM, colour = SLATE, linewidth = 0.35, alpha = 0.6) +
  scale_x_log10() +
  scale_colour_manual(values = setNames(cols, paste0("response, tau = ", taus))) +
  coord_cartesian(ylim = c(0, 103)) +
  labs(x = "Agonist concentration (nM, log scale)",
       y = "Occupancy or response (% of system max)", colour = NULL,
       title = "Increasing tau raises the ceiling and shifts EC50 left of KA",
       subtitle = "dashed = occupancy, which does not move") +
  theme(legend.position = "top", legend.text = element_text(size = 8))
```

```{r}
tau <- 9
cat(sprintf("tau = %g\n", tau))
cat(sprintf("  KA                : %.0f nM\n", KA / nM))
cat(sprintf("  EC50              : %.0f nM   <- %.0fx left of KA\n",
            ec50_obs(tau) / nM, KA / ec50_obs(tau)))
cat(sprintf("  occupancy at EC50 : %.1f%%\n", 100 * occupancy(ec50_obs(tau))))
```

Half the maximal response from **9% occupancy**. Reporting that EC~50~ as a
potency and then treating it as an affinity understates the concentration needed
at the target by tenfold.

## 3. Partial agonism is a property of the pair, not the molecule

One molecule, one affinity, three tissues.

```{r}
#| fig-height: 4
tissue <- data.frame(
  tau = c(30, 2.0, 0.4),
  lab = c("tau = 30 — high receptor density",
          "tau = 2 — intermediate",
          "tau = 0.4 — low density / poor coupling"))

df3 <- do.call(rbind, lapply(seq_len(nrow(tissue)), function(i)
  data.frame(A = A, y = response(A, tau = tissue$tau[i]), lab = tissue$lab[i])))

ggplot(df3, aes(A / nM, y, colour = lab)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = emax_obs(tissue$tau), linetype = "dotted",
             colour = SLATE, linewidth = 0.35) +
  scale_x_log10() +
  scale_colour_manual(values = setNames(c(TEAL, RUST, "#B5482F"), tissue$lab)) +
  coord_cartesian(ylim = c(0, 103)) +
  labs(x = "Agonist concentration (nM, log scale)",
       y = "Response (% of system max)", colour = NULL,
       title = "The same compound reads as full or partial depending on the tissue") +
  theme(legend.position = "top", legend.text = element_text(size = 8))
```

```{r}
for (tau in c(30, 2.0, 0.4)) {
  cat(sprintf("  tau = %-5g ceiling = %5.1f%% of system max   -> reads as %s agonist\n",
              tau, emax_obs(tau), if (emax_obs(tau) > 90) "full" else "partial"))
}
```

## 4. Schild analysis

Add a competitive antagonist. It divides the apparent agonist concentration by
the dose ratio `DR = 1 + [B]/KB`, so the curve shifts right without losing
height.

```{r}
#| fig-height: 4
KB  <- 50 * nM
tau <- 3

response_antag <- function(A, B, KB. = KB, tau. = 3)
  Em * tau. * A / (A * (tau. + 1) + KA * (1 + B / KB.))

Bs <- c(0, 10, 50, 250, 1250) * nM
lab_of <- function(B) if (B == 0) "no antagonist" else sprintf("[B] = %.0f nM", B / nM)

df4 <- do.call(rbind, lapply(Bs, function(B)
  data.frame(A = A, y = response_antag(A, B), lab = lab_of(B))))
df4$lab <- factor(df4$lab, levels = vapply(Bs, lab_of, ""))

ggplot(df4, aes(A / nM, y, colour = lab)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10() +
  scale_colour_manual(values = c(SLATE, "#C9A227", RUST, "#2E8B7A", TEAL)) +
  coord_cartesian(ylim = c(0, 103)) +
  labs(x = "Agonist concentration (nM, log scale)",
       y = "Response (% of system max)", colour = NULL,
       title = "Competitive antagonism: parallel rightward shift, maximum preserved") +
  theme(legend.position = "top", legend.text = element_text(size = 8))
```

The maximum is preserved — that is what *surmountable* means, and it is the
observation that distinguishes this mechanism from a non-competitive one.

```{r}
ec50s <- vapply(Bs, function(B) {
  E <- response_antag(A, B); A[which.min(abs(E - emax_obs(tau) / 2))]
}, numeric(1))
DR <- ec50s / ec50s[1]

cat(sprintf("%8s %9s %11s %10s\n", "[B] nM", "EC50 nM", "dose ratio", "1+[B]/KB"))
for (i in seq_along(Bs))
  cat(sprintf("%8.0f %9.2f %11.2f %10.2f\n",
              Bs[i] / nM, ec50s[i] / nM, DR[i], 1 + Bs[i] / KB))
```

Now the Schild regression: `log(DR - 1)` against `log[B]`.

```{r}
keep <- Bs > 0
x <- log10(Bs[keep]); y <- log10(DR[keep] - 1)
fit <- lm(y ~ x)
slope <- unname(coef(fit)[2]); intercept <- unname(coef(fit)[1])
pA2 <- -intercept / slope

cat(sprintf("  Schild slope : %.3f      (1.0 for simple competition)\n", slope))
cat(sprintf("  x-intercept  : %.3f = log10 KB\n", pA2))
cat(sprintf("  KB recovered : %.1f nM\n", 10^pA2 / nM))
cat(sprintf("  KB truth     : %.1f nM\n", KB / nM))
stopifnot(abs(slope - 1) < 0.02,
          isTRUE(all.equal(10^pA2, KB, tolerance = 0.02)))
```

```{r}
#| fig-height: 3.6
ggplot(data.frame(x = x, y = y), aes(x, y)) +
  geom_abline(slope = slope, intercept = intercept, colour = RUST, linewidth = 0.7) +
  geom_hline(yintercept = 0, colour = SLATE, linewidth = 0.35) +
  geom_point(size = 2.6, colour = TEAL) +
  geom_point(data = data.frame(x = pA2, y = 0), shape = 15, size = 2.8, colour = RUST) +
  annotate("text", x = pA2, y = 0.28, hjust = -0.05, size = 3.2, colour = SLATE,
           label = sprintf("log KB = %.2f\nKB = %.0f nM", pA2, 10^pA2 / nM)) +
  labs(x = "log10 [B]", y = "log10 (DR - 1)",
       title = sprintf("Schild plot — slope = %.2f is the falsification test", slope))
```

### Breaking the assumption

The slope came back at 1 because the simulation *was* competitive. Simulate a
mechanism that is not — here the antagonist also depresses the maximum, as an
insurmountable antagonist does — and the diagnostic reacts.

```{r}
response_insurm <- function(A, B, KB. = KB, tau. = 3) {
  tau_eff <- tau. / (1 + B / KB.)
  Em * tau_eff * A / (A * (tau_eff + 1) + KA)
}

ec_i <- vapply(Bs, function(B) {
  E <- response_insurm(A, B); A[which.min(abs(E - max(E) / 2))]
}, numeric(1))
DR_i <- ec_i / ec_i[1]
slope_i <- unname(coef(lm(log10(pmax(DR_i[keep] - 1, 1e-6)) ~ log10(Bs[keep])))[2])

cat(sprintf("  competitive     slope = %.3f   maximum preserved\n", slope))
cat(sprintf("  insurmountable  slope = %.3f   maximum falls with [B]\n\n", slope_i))
for (B in Bs)
  cat(sprintf("    [B] = %5.0f nM  ->  observed max %5.1f%% of system max\n",
              B / nM, max(response_insurm(A, B))))
```

A depressed maximum, and a Schild slope away from unity. Either observation
alone should stop you reporting an x-intercept as a `pKB`.

## 5. The allosteric ceiling

An allosteric modulator shifts agonist affinity by a cooperativity factor
`alpha`. Unlike competition, the shift **saturates**.

```{r}
alpha <- 10; KX <- 100 * nM

response_allo <- function(A, X, alpha. = alpha, KX. = KX, tau. = 3) {
  KA_app <- KA * (1 + X / KX.) / (1 + alpha. * X / KX.)
  Em * tau. * A / (A * (tau. + 1) + KA_app)
}

Xs <- c(0, 30, 100, 300, 1000, 10000, 100000) * nM
ec_x <- vapply(Xs, function(X) {
  E <- response_allo(A, X); A[which.min(abs(E - emax_obs(3) / 2))]
}, numeric(1))

cat(sprintf("%10s %9s %11s\n", "[X] nM", "EC50 nM", "fold shift"))
for (i in seq_along(Xs))
  cat(sprintf("%10.0f %9.2f %11.2f\n", Xs[i] / nM, ec_x[i] / nM, ec_x[1] / ec_x[i]))
cat(sprintf("\n  cooperativity alpha = %.0f  <- the shift cannot exceed this,\n", alpha))
cat("  however much modulator is added. Competitive antagonism has no such bound.\n")
```

```{r}
#| fig-height: 4
show <- c(0, 100, 1000, 100000) * nM
lab_x <- function(X) if (X == 0) "no modulator" else sprintf("[X] = %.0f nM", X / nM)
df5 <- do.call(rbind, lapply(show, function(X)
  data.frame(A = A, y = response_allo(A, X), lab = lab_x(X))))
df5$lab <- factor(df5$lab, levels = vapply(show, lab_x, ""))

ggplot(df5, aes(A / nM, y, colour = lab)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10() +
  scale_colour_manual(values = c(SLATE, "#C9A227", RUST, TEAL)) +
  coord_cartesian(ylim = c(0, 103)) +
  labs(x = "Agonist concentration (nM, log scale)",
       y = "Response (% of system max)", colour = NULL,
       title = "Positive allosteric modulation saturates at a 10-fold shift") +
  theme(legend.position = "top", legend.text = element_text(size = 8))
```

## What to take from this

```{r}
cat("EC50 is a property of a tissue, not a drug:\n")
for (tau in c(0.4, 2, 9, 50))
  cat(sprintf("  tau = %-5g -> EC50 = %6.1f nM   (affinity is %.0f nM in every row)\n",
              tau, ec50_obs(tau) / nM, KA / nM))
```
