---
title: "C-01 notebook — R"
subtitle: "How to read an FDA review"
execute:
  freeze: auto
---

The same three steps as the Python notebook: enumerate an application's document
set, sort it into a reading order, and reproduce the retrieval trap.

No network and no packages beyond base R plus `jsonlite`.

```{r}
#| message: false
library(jsonlite)

root <- normalizePath(".")
while (!file.exists(file.path(root, "resources.csv")) &&
       dirname(root) != root) root <- dirname(root)

live <- file.path(root, "data", "fda-documents.csv")
if (file.exists(live)) {
  docs   <- read.csv(live, stringsAsFactors = FALSE)
  SOURCE <- "data/fda-documents.csv — real, from `make fda`"
} else {
  fx <- fromJSON(file.path(root, "tests", "fixtures", "openfda",
                           "drug_drugsfda.json"), simplifyVector = FALSE)
  rows <- list()
  for (res in fx$results) {
    brand <- if (length(res$openfda$brand_name)) res$openfda$brand_name[[1]] else "?"
    for (sub in res$submissions) {
      for (d in sub$application_docs) {
        rows[[length(rows) + 1]] <- data.frame(
          brand = brand,
          application_number = res$application_number,
          submission = paste0(sub$submission_type, "-", sub$submission_number),
          doc_type = d$type, url = d$url, stringsAsFactors = FALSE)
      }
    }
  }
  docs   <- do.call(rbind, rows)
  SOURCE <- "tests/fixtures — SYNTHETIC (run `make fda` for real data)"
}
cat("source:", SOURCE, "\n")
cat(nrow(docs), "document rows\n")
```

## The document set

```{r}
for (app in sort(unique(docs$application_number))) {
  rows <- docs[docs$application_number == app, ]
  cat("\n", app, "  ", paste(unique(rows$brand), collapse = " / "),
      "  ", nrow(rows), "documents\n")
  for (sub in sort(unique(rows$submission))) {
    types <- sort(rows$doc_type[rows$submission == sub])
    cat("   ", formatC(sub, width = -12), paste(types, collapse = ", "), "\n")
  }
}
```

## Reading order

```{r}
rank_doc <- function(doc_type) {
  t <- tolower(ifelse(is.na(doc_type), "", doc_type))
  if (grepl("letter", t)) return(1L)
  if (grepl("label",  t)) return(2L)
  if (grepl("review", t)) return(3L)
  9L                                  # unknown sorts last, never first
}

docs$rank <- vapply(docs$doc_type, rank_doc, integer(1))
ordered   <- docs[order(docs$rank, docs$submission), ]
for (i in seq_len(min(8, nrow(ordered)))) {
  r <- ordered[i, ]
  cat(" ", r$rank, " ", formatC(r$doc_type, width = -34), " ",
      formatC(r$submission, width = -10), " ", substr(r$url, 1, 58), "\n")
}

stopifnot(rank_doc("Letter") == 1, rank_doc("Label") == 2,
          rank_doc("Review") == 3, rank_doc("REMS") == 9,
          rank_doc("") == 9, rank_doc(NA) == 9)
cat("\nreading order: unknown types sort last — OK\n")
```

## The trap

```{r}
fx <- fromJSON(file.path(root, "tests", "fixtures", "openfda",
                         "drug_drugsfda.json"), simplifyVector = FALSE)
WANTED <- "LUMAKRAS"

cat("querying by generic name alone returns", length(fx$results),
    "applications:\n\n")
for (res in fx$results) {
  brand <- if (length(res$openfda$brand_name)) res$openfda$brand_name[[1]] else "?"
  ndocs <- sum(vapply(res$submissions,
                      function(s) length(s$application_docs), integer(1)))
  verdict <- if (grepl(WANTED, toupper(brand), fixed = TRUE))
    "the product we meant" else
    "DIFFERENT PRODUCT — same moiety, different formulation"
  cat(" ", formatC(res$application_number, width = -12),
      formatC(brand, width = -16), ndocs, "docs  ", verdict, "\n")
}
```

```{r}
assert_brand <- function(result, expected) {
  brands <- toupper(unlist(result$openfda$brand_name))
  if (!any(grepl(expected, brands, fixed = TRUE)))
    stop(sprintf("expected '%s', got %s", expected,
                 paste(brands, collapse = ", ")))
  invisible(result)
}

assert_brand(fx$results[[1]], "LUMAKRAS")
cat("matching application: passes\n")

res <- try(assert_brand(fx$results[[2]], "LUMAKRAS"), silent = TRUE)
cat("impostor application: raises ->",
    sub("^Error[^:]*: ", "", conditionMessage(attr(res, "condition"))), "\n")
```

## What you should take away

The package is a set, the disciplines own different claims, and the retrieval
step is where the errors enter. Assert on identity before you read a parameter.
